---
title: "README"
output: github_document
---

[![Travis build status](https://travis-ci.org/pierredenelle/Bioregionalization.svg?branch=master)](https://travis-ci.org/pierredenelle/Bioregionalization)

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`Bioregionalization` is a package designed to compare several methods of clustering based on multivariate or network approaches.

## 1 Installation

```r
devtools::install_github("pierredenelle/Bioregionalization", build_vignettes = TRUE)
```

## 2 Tutorial

A tutorial vignette showing the main steps of the processing can be visualised with the following command line:

```{r}
vignette("tutorial", package = "Bioregionalization")
```

The .pdf of the vignette can also be accessed **[here](https://github.com/pierredenelle/Bioregionalization/blob/master/documentation/tutorial_documentation.pdf)**.

## 3 Dependencies
`Bioregionalization` depends on `dplyr`, `ecodist`, `reshape2`, `DescTools`,
`ade4`, `cluster`, `sf`, `ggplot2`, `Rcpp`, `SMUT` and `igraph`.
