---
title: "1. Installation of the binary files"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: false
  html_document:
    toc: true
    toc_float:
    collapsed: false
    smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. Installation of the binary files}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: '`r system.file("REFERENCES.bib", package="bioregion")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 8)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library(bioregion)
  })
})

options(tinytex.verbose = TRUE)

```

Some functions (listed below) or at least part of them require binary
files to run.

-   [netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html)
-   [netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html)
    (Cpp version)
-   [netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html)

The function
[install_binaries](https://bioRgeo.github.io/bioregion/reference/install_binaries.html)
downloads and unzips the `bin` folder needed to run these functions. It
also checks if the files have the permissions to be executed as
programs. It finally tests if the binary files are running properly.

### Run install_binaries()

The function
[install_binaries](https://bioRgeo.github.io/bioregion/reference/install_binaries.html)
should be run prior to using
[netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html),
the Cpp version of
[netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html)
and
[netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html)
as follows.

```{r install_binaries, results='markup', message=TRUE, warning=TRUE, echo=TRUE}
install_binaries(binpath = "tempdir" ,
                 download_only =  FALSE,
                 infomap_version = c("2.1.0", 
                                     "2.6.0",
                                     "2.7.1",
                                     "2.8.0"))
```

The function has three parameters. `binpath` indicating the path to the
folder that will host the `bin` folder containing the binary files. By
default, the binary files are installed in R's temporary directory
(`binpath = "tempdir"`). In this case the `bin` folder will be
automatically removed at the end of the R session. Alternatively, the
binary files can be installed in the bioregion's package folder
(`binpath = "pkgfolder"`). **Any other folder can be chosen but in any
case PLEASE MAKE SURE to update the `binpath` argument in
[netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html),
[netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html)
and
[netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html)
accordingly.** Note that in this case, and only in this case,
`download_only` can be set to `TRUE` (only the first step will be
executed), but you must ensure that the files have the necessary
permissions to be executed as programs otherwise.

The third parameter `infomap_version` indicating the Infomap version(s)
to install. Only the Infomap version 2.1.0, 2.6.0, 2.7.1 and 2.8.0 are
available for now.

The installation of the binary files is divided into six steps:

1. Download bin.zip
2. Unzip bin.zip in the binpath
3. Check the permissions, try to change them automatically or propose an iterative process to change them manually
4. Test that Infomap are running properly
5. Test that Louvain are running properly
6. Test that OSLOM are running properly

The function is designed to help you change the permissions during the third 
step. Subsequently, the binary files of Infomap, Louvain, and OSLOM are tested, 
and a file `check.txt` is added to each folder `bin/INFOMAP/X.X.X`, `bin/LOUVAIN`
and `bin/OSLOM` to inform the package that [netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html),
[netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html)
and
[netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html) 
can be used without any issues.

However, it may happen (particularly on macOS) that the function fails to 
automatically or manually assist you in changing the permissions. In this case, 
set the argument `dowload_only` to `TRUE` to download `bin.zip` into a folder 
`myfolder` of your choice (different from `tempdir` and `pkgfolder`) so that you
can manually change the permissions yourself. Note that you will be on your own;
the package will not be able to test the binary files before running [netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html),
[netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html)
or
[netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html) 
(i.e., no `check.txt` will be added to the folders...). To bypass the automatic
check while using these functions, use the argument `check_install = FALSE`, but 
without any guarantee of success.

```{r install_binaries_download_only, eval = FALSE}
install_binaries(binpath = "myfolder" ,
                 download_only =  TRUE,
                 infomap_version = c("2.1.0", 
                                     "2.6.0",
                                     "2.7.1",
                                     "2.8.0"))

comat <- matrix(sample(1000, 50), 5, 10)
rownames(comat) <- paste0("Site", 1:5)
colnames(comat) <- paste0("Species", 1:10)

net <- similarity(comat, metric = "Simpson")
com <- netclu_infomap(net,
                      binpath = "myfolder",
                      check_install = TRUE)
```

### Known issues

-   The OpenMP versions of Infomap require `libomp-dev` on Ubuntu
    (`sudo apt-get install libomp-dev`) and `libomp` on macOS (install
    [Homebrew](https://brew.sh) and run `brew install libomp`).
