---
title: "1. Installation of the binary files"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: false
  html_document:
    toc: true
    toc_float:
    collapsed: false
    smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. Installation of the binary files}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: '`r system.file("REFERENCES.bib", package="bioregion")`'
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 8)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library(bioregion)
  })
})

options(tinytex.verbose = TRUE)

```

Some functions (listed below) or at least part of them require binary
files to run. 

* [netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html)
* [netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html) (Cpp version)
* [netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html)

The function [install_binaries](https://bioRgeo.github.io/bioregion/reference/install_binaries.html)
downloads and unzips the `bin` folder needed to run these functions. It also 
checks if the files have the permissions to be executed as programs. It finally 
tests if the binary files are running properly.

### Run install_binaries()

The function [install_binaries](https://bioRgeo.github.io/bioregion/reference/install_binaries.html) 
should be run prior to using 
[netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html),
the Cpp version of [netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html) 
and [netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html) 
as follows.

```{r install_binaries_show, eval=FALSE}
install_binaries(binpath = "tempdir" ,
                 download_only =  FALSE,
                 infomap_version = c("2.1.0", 
                                     "2.6.0",
                                     "2.7.1",
                                     "2.8.0"))
```

The function has three parameters. `binpath` indicating the path to the folder 
that will host the `bin` folder containing the binary files. By default, the 
binary files are installed in R's temporary directory (`binpath = "tempdir"`). 
In this case the `bin` folder will be automatically removed at the end of the R 
session. Alternatively, the binary files can be installed in the bioregion's 
package folder (`binpath = "pkgfolder"`). **Any other folder can be chosen but 
in any case PLEASE MAKE SURE to update the binpath argument in 
[netclu_infomap](https://bioRgeo.github.io/bioregion/reference/netclu_infomap.html), 
[netclu_louvain](https://bioRgeo.github.io/bioregion/reference/netclu_louvain.html)
and [netclu_oslom](https://bioRgeo.github.io/bioregion/reference/netclu_oslom.html)
accordingly.** Note that in this case, and only in this case, `download_only` 
can be set to `TRUE` (only the first step will be executed), but you must ensure
that the files have the necessary permissions to be executed as programs 
otherwise.

The third parameter `infomap_version` indicating the Infomap 
version(s) to install. Only the Infomap version 2.1.0, 2.6.0, 2.7.1 and 
2.8.0 are available for now.

The installation of the binary files is divided into six steps:

#### 1. Download bin.zip 

#### 2. Unzip bin.zip in the binpath 

#### 3. Check the permissions, try to change them automatically or propose an iterative process to change them manually

#### 4-> 6. Test that Infomap, Louvain and OSLOM are running properly 

```{r install_binaries}
install_binaries(binpath = "tempdir" ,
                 download_only =  FALSE,
                 infomap_version = c("2.1.0", 
                                     "2.6.0",
                                     "2.7.1",
                                     "2.8.0"))
```

###  Known issues

* The OpenMP versions of Infomap require `libomp-dev` on Ubuntu 
(`sudo apt-get install libomp-dev`) and `libomp` on macOS (install [Homebrew](https://brew.sh) 
and run `brew install libomp`).








