---
title: "1. Matrix and network formats"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{1. Matrix and network formats}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 8)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library(bioRgeo)
  })
})

options(tinytex.verbose = TRUE)

```

test

Some functions (listed below) or at least part of them require binary files to run. 

* [netclu_infomap](https://biorgeo.github.io/bioRgeo/reference/netclu_infomap.html)
* [netclu_louvain](https://biorgeo.github.io/bioRgeo/reference/netclu_louvain.html)
* [netclu_oslom](https://biorgeo.github.io/bioRgeo/reference/netclu_oslom.html)

The function \code{install_binaries} downloads and unzips the bin folder needed to run these functions. It also checks if the files have the permissions to be executed as programs. It finally tests if the executable are running properly.

``` {r install_binaries}
#install_binaries()
```

https://brew.sh

brew install libomp

```{r loading_virtual_data}
data(vegedf)
head(vegedf)
```




