---
title: "2. Matrix and network formats"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
bibliography: '`r system.file("REFERENCES.bib", package="bioregion")`' 
csl: journal-of-biogeography.csl    
vignette: >
  %\VignetteIndexEntry{2. Matrix and network formats}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 8)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library(bioregion)
  })
})

options(tinytex.verbose = TRUE)

```

## Load data

The `bioregion`'s package contains as example dataset the spatial distribution of 
Mediterranean vegetation. This dataset has been analyzed in [this article](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.4718) and contains
the abundance of 3,697 species in 715 sites. This dataset is composed of three 
files, [vegedf](https://bioRgeo.github.io/bioregion/reference/vegedf.html) a 
`data.frame` with 460,878 rows and 3 columns (Site, Species and Abundance), 

```{r loading_vegedf}
data(vegedf)
head(vegedf)
dim(vegedf)
sum(!duplicated(vegedf[,1]))
sum(!duplicated(vegedf[,2]))
```

[vegemat](https://bioRgeo.github.io/bioregion/reference/vegemat.html) 
a co-occurrence `matrix` containing the same information gathered in a `matrix`
with 715 rows and 3,697 columns,

```{r loading_vegemat}
data(vegemat)
vegemat[1:10,1:10]
dim(vegemat)
```

and [vegesf](https://bioRgeo.github.io/bioregion/reference/vegesf.html) a spatial
object containing the geometry of the 715 sites. 

## From matrix to network

The function [mat_to_net](https://bioRgeo.github.io/bioregion/reference/mat_to_net.html) 
transforms a co-occurrence `matrix` such as `vegemat` into a network represented 
by a `data.frame` (such as `vegedf` in this case). If `weight = TRUE` a third 
column is added with the values contained in the `matrix`.  

``` {r mat_to_net_1}
net <- mat_to_net(vegemat, weight = TRUE, remove_zeroes = FALSE)
```

In line with the network format, the two first columns are named `Node1` and 
`Node2` by default.

```{r print_net_1}
head(net)
dim(net)
```

If `remove_zeroes = TRUE` the pairs of nodes with a weight equal to 0 will 
be removed from the output.

``` {r mat_to_net_2}
net <- mat_to_net(vegemat, weight = TRUE, remove_zeroes = TRUE)
```

```{r print_net_2}
head(net)
dim(net)
```

## From network to matrix

The function [net_to_mat](https://bioRgeo.github.io/bioregion/reference/net_to_mat.html)
does the opposite. It transforms a network represented by a two- or a three-columns
`data.frame` (such as `vegedf`) into a co-occurrence `matrix` (such as `vegemat` 
in this case). 

``` {r net_to_mat_1}
mat <- net_to_mat(vegedf, weight = TRUE, squared = FALSE, symmetrical = FALSE, missing_value = 0)
```

```{r print_mat_1}
mat[1:5,1:5]
dim(mat)
```

If `squared = TRUE` a squared matrix will be generated, the rownames and colnames 
will correspond to the concatenation without duplicates of the two first 
columns of the `data.frame`. 

``` {r net_to_mat_2}
mat <- net_to_mat(vegedf, weight = TRUE, squared = TRUE, symmetrical = FALSE, missing_value = 0)
```

```{r print_mat_2}
mat[1:5,1:5]
dim(mat)
```

The argument `missing_value` defines the value to assign to the pairs of nodes 
not present in the input network. The default value is 0 but any other numeric
value can be used.

``` {r net_to_mat_3}
temp <- data.frame(Site=c("35","36","36","38","39"), Species=c("36","35","37","37","39"), Abundance=c(1,2,3,4,0))
net <- rbind(temp,vegedf)
mat <- net_to_mat(net, weight = TRUE, squared = TRUE, symmetrical = FALSE, missing_value = -1)
```

```{r print_mat_3}
mat[1:5,1:5]
```

Finally, if `squared = TRUE` it is possible to get a symmetrical matrix as output
(`symmetrical = TRUE`). In this case the resulting squared matrix will be 
symmetrical, except for the symmetrical pairs of nodes already present in the
input network (35 <-> 36) in the example below.

``` {r net_to_mat_4}
mat <- net_to_mat(net, weight = TRUE, squared = TRUE, symmetrical = TRUE, missing_value = 0)
```

```{r print_mat_4}
mat[1:5,1:5]
```
