---
title: "3. Pairwise similarity/dissimilarity metrics"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
bibliography: '`r system.file("REFERENCES.bib", package="bioregion")`' 
csl: journal-of-biogeography.csl    
vignette: >
  %\VignetteIndexEntry{3. Pairwise similarity/dissimilarity metrics}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 8)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library(bioregion)
  })
})

options(tinytex.verbose = TRUE)

```

```{r loading_vegemat, include=FALSE}
data(vegemat)
```

The function [similarity](https://bioRgeo.github.io/bioregion/reference/similarity.html) 
compute well-known and customized pairwise similarity metrics based on a 
co-occurrence `matrix` such as `vegemat`. In the example below the Simpson similarity
index is computed between each pair of sites.

``` {r similarity_1}
sim <- similarity(vegemat, metric = "Simpson", formula = NULL, method = "prodmat")
sim[1:10,]
```

The resulting `data.frame` is stored in a 
`bioregion.pairwise.metric` object containing Simpson similarity metric between each 
pair of sites. The function [similarity](https://bioRgeo.github.io/bioregion/reference/similarity.html) 
can handle three types of metrics: the metrics based on `abc`, the metrics based 
on `ABC` and one metric based on the Euclidean distance. 

The first kind of metrics such as Jaccard, the turnover component of Jaccard, 
Simpson or Sorensen are based on presence data with `a` the number of species 
shared by a pair of 
sites, `b` species only present in the first site and `c` species only present 
in the second site. Two methods can be used to compute the `abc` based metrics. 
The first method is based on a matrix product (performed with the [tcrossprod](https://www.rdocumentation.org/packages/Matrix/versions/1.5-1/topics/matrix-products) 
function from the R package [Matrix](https://www.rdocumentation.org/packages/Matrix/versions/1.5-1)). 
The method is fast but is greedy in memory... The second method is 
based on a [three loops function](https://github.com/bioregion/bioregion/blob/master/src/abc.cpp) 
coded in C++ and largely inspired by the [bcdist](https://www.rdocumentation.org/packages/ecodist/versions/2.0.9/topics/bcdist) 
function from the R package [ecodist](https://www.rdocumentation.org/packages/ecodist/versions/2.0.9) 
(version 2.0.7). It is less efficient than the matrix product but allows to handle 
co-occurrence matrix with a large number of sites and/or species.

The second kind of metrics such as Bray-Curtis and the turnover component of 
Bray-Curtis are based on abundance data with `A` the sum of the lesser values for 
common species shared by a pair of sites. `B` and `C` are the total number of 
specimens counted at both sites minus `A`. Only three loops function is 
available for the `ABC` based metrics. 

The main advantage of the [similarity](https://bioRgeo.github.io/bioregion/reference/similarity.html) 
function is to compute and return several metrics, to allow the computation of
customized metric with the `formula` argument and to include the possibility of
returning `a`, `b` and `c` and/or `A`, `B` and `C`. This feature is particularly 
interesting to compute similarity metrics on large co-occurrence matrix. 

``` {r similarity_2}
sim <- similarity(vegemat, metric = c("abc","ABC","Simpson","Bray"), formula =c("(b + c) / (a + b + c)", "(B + C) / (2*A + B + C)"))
sim[1:10,]
```

The [dissimilarity](https://bioRgeo.github.io/bioregion/reference/dissimilarity.html) 
function is very similar, with the sole exception that it computes the 
dissimilarity version of the available metrics. The functions [dissimilarity_to_similarity](https://bioRgeo.github.io/bioregion/reference/dissimilarity_to_similarity.html) and [similarity_to_dissimilarity](https://bioRgeo.github.io/bioregion/reference/dissimilarity_to_similarity.html)
can be used to switch between similarity and dissimilarity metrics.

``` {r dissimilarity}
sim <- similarity(vegemat, metric = c("abc","Simpson"), formula = "(b + c) / (a + b + c)")
sim[1:10,]
dissim1 <- dissimilarity(vegemat, metric = c("abc","Simpson"), formula = "(b + c) / (a + b + c)")
dissim1[1:10,]
dissim2 <- similarity_to_dissimilarity(sim)
dissim2[1:10,]
```


