---
title: "5.2 Summary metrics"
author: "Pierre Denelle, Boris Leroy and Maxime Lenormand"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
bibliography: '`r system.file("REFERENCES.bib", package="bioregion")`' 
csl: style_citation.csl    
vignette: >
  %\VignetteIndexEntry{5.2 Summary metrics}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 6)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library("bioregion")
    library("dplyr")
    library("ggplot2")
    library("sf")
  })
})

options(tinytex.verbose = TRUE)
```

In this vignette, we aim to evaluate the contribution of individual species and 
sites to each bioregion using the function `site_species_metrics()`. We also show 
how various metrics at the bioregion level can be computed, in particular with 
the function `bioregion_metrics()`.

# 1. Data  

We use the vegetation dataset included in the `bioregion`.

```{r}
data("vegedf")
data("vegemat")

# Calculation of (dis)similarity matrices
vegedissim <- dissimilarity(vegemat, metric = c("Simpson"))
vegesim <- dissimilarity_to_similarity(vegedissim)
```

# 2. Bioregionalization

We use the same three bioregionalization algorithms as in the
[visualization vignette](https://biorgeo.github.io/bioregion/articles/a5_visualization.html),
i.e., non-hierarchical, hierarchical, and network bioregionalizations. In 
addition, we include a network bioregionalization algorithm based on a bipartite 
network, which assigns a bioregion (or cluster) to both sites and species. We 
chose three bioregions for the non-hierarchical and hierarchical 
bioregionalizations.
<br>

```{r}
# Non hierarchical bioregionalization
vege_nhclu <- nhclu_kmeans(vegedissim, 
                           n_clust = 3, 
                           index = "Simpson",
                           seed = 1)
vege_nhclu$cluster_info 

# Hierarchical bioregionalization
set.seed(1)
vege_hclu <- hclu_hierarclust(dissimilarity = vegedissim,
                              index = "Simpson",
                              method = "average", 
                              n_clust = 3,
                              optimal_tree_method = "best",
                              verbose = FALSE)
vege_hclu$cluster_info

# Network bioregionalization
set.seed(1)
vege_netclu <- netclu_walktrap(vegesim,
                               index = "Simpson")
vege_netclu$cluster_info 

# Bipartite network bioregionalization
install_binaries(verbose = FALSE)
vege_netclubip <- netclu_infomap(vegedf,
                                 seed = 1, 
                                 bipartite = TRUE)
vege_netclubip$cluster_info

```

# 3. Species-to-bioregion & site-to-bioregion metrics

Several metrics evaluating the contribution of a species or a site to each 
bioregion can be computed and further aggregated to assess the relationship 
between a species or a site and the bioregionalization.

## 3.1 Species-to-bioregion metrics

Species-to-bioregion metrics are based on a bioregionalization containing $K$ 
bioregions (`vege_nhclu` computed above for example) and its associated 
site–species co-occurrence matrix
[vegemat](https://biorgeo.github.io/bioregion/reference/vegemat.html) to
compute three core terms, from which occurrence-based indices are derived.

- $n_{sb}$ is the number of sites belonging to bioregion $b$ in which species $s$ is present.  
- $n_s$ is the total number of sites in which species $s$ is present.  
- $n_b$ is the total number of sites belonging to bioregion $b$.

<center>
  <img align="bottom" 
   width="70%" height="70%" 
   src="../man/figures/core_terms.svg"
   alt="Illustration of core terms.">
</center> 

```{r}
nsb <- site_species_metrics(bioregionalization = vege_nhclu,
                            bioregion_indices = "CoreTerms",
                            bioregionalization_indices = NULL,
                            data_type = "occurrence",
                            node_type = "site",
                            comat = vegemat,
                            similarity = NULL,
                            index = NULL,
                            verbose = FALSE)

nsb$species_bioregions[1:10,]

```

These core terms and associated indices are computed when 
`data_type = "occurrence"` (or `data_type = "both"`) and 
`node_type = "site"` (or `node_type = "both"`). `node_type = "site"` means that 
a bioregion is assigned to each site in `vege_nhclu`. 

Several metrics can be derived from these core terms.

### Specificity (occurrence)

The specificity $A_{sb}$ of species $s$ for bioregion $b$ [@Caceres2009] is 
defined as

$$A_{sb} = \frac{n_{sb}}{n_s}$$

and measures the fraction of occurrences of species $s$ that belong to 
bioregion $b$. It therefore reflects the uniqueness of a species to a particular 
bioregion.

### NSpecificity (occurrence)

A normalized version that accounts for the size of each bioregion is also 
available, as defined in [@Caceres2009]:

$$\bar{A}_{sb} = \frac{n_{sb}/n_b}{\sum_{k=1}^K n_{sk}/n_k}$$

It corresponds to a normalized specificity value that adjusts for differences 
in bioregion size.

### Fidelity (occurrence)

The fidelity $B_{sb}$ of species $s$ for bioregion $b$ [@Caceres2009] is 
defined as

$$B_{sb} = \frac{n_{sb}}{n_b}$$

and measures the fraction of sites in bioregion $b$ where species $s$ is 
present. It therefore reflects the frequency of occurrence of a species 
within a bioregion.

### IndVal (occurrence)

The indicator value $IndVal_{sb}$ of species $s$ for bioregion $b$ can be 
defined as the product of specificity and fidelity [@Caceres2009]:

$$IndVal_{sb} = A_{sb} \times B_{sb}$$

This index quantifies the strength of association between a species and a 
bioregion by combining its specificity (uniqueness to that bioregion) and 
fidelity (consistency of occurrence within that bioregion). High IndVal 
values identify species that are both frequent and restricted to a single 
bioregion, making them good indicators of that region.

### NIndVal (occurrence)

A normalized version of the indicator value is also available:

$$\bar{IndVal}_{sb} = \bar{A}_{sb} \times B_{sb}$$

This normalization adjusts for differences in bioregion size, allowing more 
comparable indicator values across regions with unequal sampling effort or 
extent.

### Rho (occurrence)

The contribution index $\rho$ can also be calculated following 
[@Lenormand2019]:

$$\rho_{sb} = \frac{n_{sb} - n_s\frac{n_b}{n}}{\sqrt{\frac{n_b(n - n_b)}{n - 1} 
\frac{n_s}{n}(1 - \frac{n_s}{n}) }}$$

This index measures the deviation between the observed number of occurrences of 
species $s$ in bioregion $b$ and the expected value under random association, 
providing a standardized measure of contribution to the bioregional structure.

```{r}
nsb <- site_species_metrics(bioregionalization = vege_nhclu,
                            bioregion_indices = c("Specificity", "NSpecificity",
                                                  "Fidelity",
                                                  "IndVal", "NIndVal",
                                                  "Rho",
                                                  "CoreTerms"),
                            bioregionalization_indices = NULL,
                            data_type = "occurrence",
                            node_type = "site",
                            comat = vegemat,
                            similarity = NULL,
                            index = NULL,
                            verbose = FALSE)

nsb$species_bioregions[1:10,]

```

The abundance version of these metrics can also be computed when 
`data_type = "abundance"` (or `data_type = "both"`). In this case the core terms 
and associated indices are:

- $w_{sb}$ is the sum of abundances of species **s** in sites of bioregion **b**. 
- $w_s$ is the total abundance of species **s**.  
- $w_b$ is the total abundance of all species present in sites of bioregion **b**.


```{r}
wsb <- site_species_metrics(bioregionalization = vege_nhclu,
                            bioregion_indices = "CoreTerms",
                            bioregionalization_indices = NULL,
                            data_type = "abundance",
                            node_type = "site",
                            comat = vegemat,
                            similarity = NULL,
                            index = NULL,
                            verbose = FALSE)

wsb$species_bioregions[1:10,]

```

### Specificity (abundance)

$$A_{sb} = \frac{w_{sb}}{w_s}$$

### NSpecificity (abundance)

$$\bar{A}_{sb} = \frac{w_{sb}/n_b}{\sum_{k=1}^K w_{sk}/n_k}$$

### Fidelity (abundance)

$$B_{sb} = \frac{w_{sb}}{w_b}$$

### IndVal (abundance)

$$IndVal_{sb} = A_{sb} \times \frac{n_{sb}}{n_b}$$
Note that the fidelity based on occurrence is used here [@Caceres2009].

### NIndVal (abundance)

$$\bar{IndVal}_{sb} = \bar{A}_{sb} \times \frac{n_{sb}}{n_b}$$

Note that the fidelity based on occurrence is used here [@Caceres2009].

### Rho (abundance)

$$\rho_{sb} = \frac{\mu_{sb} - \mu_s}{\sqrt{\left(\frac{n - n_b}{n-1}\right) \left(\frac{{\sigma_s}^2}{n_b}\right)}}$$
where 

- $\mu_{sb} = \frac{w_{sb}}{n_b}$ the average abundance of species $s$ in 
bioregion $b$ (as in *NSpecificity* and *NIndVal*)
- $\mu_s = \frac{w_s}{n}$ the average abundance of species $s$
- $\sigma_s$ the associated standard deviation.


```{r}
wsb <- site_species_metrics(bioregionalization = vege_nhclu,
                            bioregion_indices = c("Specificity", "NSpecificity",
                                                  "Fidelity",
                                                  "IndVal", "NIndVal",
                                                  "Rho",
                                                  "CoreTerms"),
                            bioregionalization_indices = NULL,
                            data_type = "abundance",
                            node_type = "site",
                            comat = vegemat,
                            similarity = NULL,
                            index = NULL,
                            verbose = FALSE)

wsb$species_bioregions[1:10,]

```

## 3.2 Species-to-bioregionalization metrics

Based on all these indices, it is also possible to derive aggregated indices.  
For now, only the participation coefficient $P_s$ of a species $s$ to the 
bioregionalization as described in [@Denelle2020] is proposed, available in 
both its occurrence and abundance versions.

### P (occurrence)

$$
P_s =  1 - \sum_{k=1}^K \left(\frac{n_{sk}}{n_s}\right)^2
$$

### P (abundance)

$$
P_s =  1 - \sum_{k=1}^K \left(\frac{w_{sk}}{w_s}\right)^2
$$

These indices measure how evenly a species is distributed among bioregions. 
There are ranging from 0 to 1. Values close to 0 indicate that the species is 
largely restricted to a single bioregion, while values close to 1 indicate that 
the species is evenly distributed across multiple bioregions.

```{r}
ps <- site_species_metrics(bioregionalization = vege_nhclu,
                           bioregion_indices = NULL,
                            bioregionalization_indices = "P",
                            data_type = "both",
                            node_type = "site",
                            comat = vegemat,
                            similarity = NULL,
                            index = NULL,
                            verbose = FALSE)

ps$species_bioregionalization[1:10,]

```

## 3.3 Site-to-species cluster & site-to-species clustering metrics

Dual metrics can also be computed when a cluster or bioregion has been assigned  
to each species in a bioregionalization (as is the case for `vege_netclubip`  
computed above). In this case, setting `node_type = "species"` (or  
`node_type = "both"`) allows computing site-to-species cluster and/or  
site-to-species clustering metrics. These metrics are derived from the same  
principles as above, based on three core terms:

- $n_{gc}$ is the number of species belonging to cluster $c$ that are present in site $g$.  
- $n_g$ is the total number of species present in site $g$.  
- $n_c$ is the total number of species belonging to cluster $c$.  

and their abundance counterparts.

```{r}
gc <- site_species_metrics(bioregionalization = vege_netclubip,
                            bioregion_indices = c("Specificity", "NSpecificity",
                                                  "Fidelity",
                                                  "IndVal", "NIndVal",
                                                  "Rho",
                                                  "CoreTerms"),
                            bioregionalization_indices = "P",
                            data_type = "both",
                            node_type = "species",
                            comat = vegemat,
                            similarity = NULL,
                            index = NULL,
                            verbose = FALSE)

gc$site_clusters[1:10,]
gc$site_clustering[1:10,]

```

## 3.4 Site-to-bioregion metrics

A final type of metrics can be computed when a bioregionalization containing  
$K$ bioregions (for example, `vege_nhclu` computed above) and its associated  
site similarity matrix (`vegesim`) are available.

These metrics include the average similarity of each site to the sites of  
each bioregion ($MeanSim$) and the associated standard deviation ($SdSim$).  
When computing the average similarity, the focal site itself is not included  
in the calculation for its own bioregion.

### MeanSim

Let $g$ be a site and $b$ a bioregion with sites $g' \in b$, then:

$$MeanSim_{gb} = \frac{1}{n_b - \delta_{g \in b}} \sum_{g' \in b, g' \neq g} sim_{gg'}$$
where $sim_{gg'}$ is the similarity between sites $g$ and $g'$, $n_b$ is the 
number of sites in bioregion $b$, and $\delta_{g \in b}$ is 1 if site $g$ belongs 
to bioregion $b$ (to exclude itself), 0 otherwise.

### SdSim

The standard deviation of similarities of site $g$ to bioregion $b$ is:

$$SdSim_{gb} = \sqrt{\frac{1}{n_b - 1 - \delta_{g \in b}} \sum_{g' \in b, g' \neq g} \left( sim_{gg'} - MeanSim_{gb} \right)^2}$$
where $sim_{gg'}$ is the similarity between sites $g$ and $g'$, $n_b$ is the 
number of sites in bioregion $b$, and $\delta_{g \in b}$ is 1 if site $g$ 
belongs to bioregion $b$ (to exclude itself), 0 otherwise.


## 3.5 Site-to-bioregionalization metrics

Based on $MeanSim$, it is possible to derive aggregated metrics that assess  
how well a site fits within its assigned bioregion relative to others.  

For now, only the Silhouette index [@Rousseeuw1987] is proposed.

### Silhouette

The Silhouette index for a site $g$ is defined as:

$$Silhouette_g = \frac{a_g - b_g}{\max(a_g, b_g)}$$

where:

- $a_g$ is the average similarity of site $g$ to all other sites in its own bioregion,  
- $b_g$ is the average similarity of site $g$ to all sites belonging to the nearest bioregion.

This index reflects how strongly a site is associated with its assigned bioregion 
relative to the most similar alternative bioregion, ranging from -1, when the site 
may be misassigned (i.e., more similar to another bioregion than its own), to 1, 
when the site is well matched to its own bioregion, and around 0 when the site 
lies near the boundary between bioregions.

# 4. Bioregion metrics & spatial coherence

For each bioregion, we calculate the number of sites it contains and the number 
of species present in those sites. The number and proportion of endemic species 
are also computed. Endemic species are defined as those occurring only in sites 
assigned to a particular bioregion.

```{r}
bioregion_summary <- bioregion_metrics(bioregionalization = vege_nhclu,
                                       comat = vegemat)
bioregion_summary
```

We use the metric of spatial coherence as in [@Divisek2016], except that we
replace the number of pixels per bioregion with the area of each coherent part.

The spatial coherence is expressed in percentage, and has the following
formula:

$$SC_j = 100 \times \frac{LargestPatch_j}{Area_j}$$

where $j$ is a bioregion.

Here is an example with the vegetation dataset.

```{r}
# Spatial coherence
vegedissim <- dissimilarity(vegemat)
hclu <- nhclu_kmeans(dissimilarity = vegedissim, n_clust = 4)
vegemap <- map_bioregions(hclu, vegesf, write_clusters = TRUE, plot = FALSE)

bioregion_metrics(bioregionalization = hclu, comat = vegemat, map = vegemap,
col_bioregion = 2) 
```

The bioregion 4 is almost constituted of one homogeneous block, which is why 
the spatial coherence is very close to 100 %.

```{r}
ggplot(vegemap) +
  geom_sf(aes(fill = as.factor(K_4))) +
  scale_fill_viridis_d("Bioregion") +
  theme_bw() +
  theme(legend.position = "bottom")
```


# 5. References
