---
title: "5. Visualization"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{5. Visualization}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 6)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library("bioregion")
    library("ggplot2")
    library("dplyr")
    library("rnaturalearth")
    library("sf")
    library("tidyr")
  })
})

options(tinytex.verbose = TRUE)
```

In this vignette, we aim at illustrating how to plot the bioregions identified
with the diferent algorithms available in `bioregion`.
<br>
Using one of the dataset coming along with `bioregion`, we show three strategies
to plot your results.

# Data  
For this vignette, we rely on the dataset describing the distribution of fresh
water fish in Europe. We first load the matrix format of this dataset, computes
the dissimilarity matrix out of it and also load the data.frame format of the
data.

```{r}
data("fishmat")
fishdissim <- dissimilarity(fishmat, metric = "all")
data("fishdf")
```

Since we aim at plotting the result, we also need the object `fishsf` linking
each site of the dataset to a geometry.

```{r}
data("fishsf")
```

We also import the world coastlines, available from the `rnaturalearth`
R package.

```{r}
world <- rnaturalearth::ne_coastline(returnclass = "sf")
```

# Plots  
In this section, we show three ways to plot your results.
<br>

## map_clusters()  
The first possibility is to use the function `map_clusters()` from the package.
This function can directly provide a plot of each site colored according to
the cluster they belong to.
<br>
Let's take an example with a K-means clustering, with a number of clusters set
to 5.

```{r}
fish_nhclu_kmeans <- nhclu_kmeans(fishdissim, n_clust = 5, index = "Simpson")
```

`map_clusters()` function can now simply takes the object `fish_nhclu_kmeans`,
which is of `bioregion.clusters` class, and the spatial distribution of sites,
stored in `fishsf`.

```{r}
map_clusters(fish_nhclu_kmeans, geometry = fishsf, plot = TRUE)
```

## Custom plot  
If you want to customize yourself the plot and not simply rely on the default
option, `map_clusters()` gives you the possibility to extract each site as well
as its geometry and cluster number.  
For this purpose, you can set the arguments like in the chunk below:

```{r}
custom <- map_clusters(fish_nhclu_kmeans, geometry = fishsf,
                       write_clusters = TRUE, plot = FALSE)
custom

# Crop world coastlines to the extent of the sf object of interest
europe <- sf::st_crop(world, sf::st_bbox(custom))

# Plot
ggplot(custom) +
  geom_sf(aes(fill = K_5), show.legend = FALSE) +
  geom_sf(data = europe) +
  scale_fill_viridis_d() +
  labs(title = "Kmeans with 5 clusters") +
  theme_void()
```

## Plot with facets  
Finally, you can be interested in plotting several bioregionalizations at once.
For this purpose, we can build a single `data.frame` gathering all the
bioregions obtained from distinct algorithms and then take advantage of the
`facets` implemented in `ggplot2`.
<br>

We first compute a few more bioregionalization on the same dataset using
other algorithms.
<br>

```{r}
fish_nhclu_pam <- nhclu_pam(fishdissim, n_clust = 5, index = "Simpson")
fish_hclu_hierarclust <- hclu_hierarclust(dissimilarity = fishdissim,
                                          n_clust = 5)
fish_netclu_greedy <- netclu_greedy(fishdf)
```

We can now make a single `data.frame` with an extra-column indicating the
algorithm used.

```{r}
fish_kmeans <- fish_nhclu_kmeans$clusters
colnames(fish_kmeans)<- c("ID", "NHCLU_KMEANS")
fish_pam <- fish_nhclu_pam$clusters
colnames(fish_pam)<- c("ID", "NHCLU_PAM")
fish_hieraclust <- fish_hclu_hierarclust$clusters
colnames(fish_hieraclust)<- c("ID", "HCLU_HIERARCLUST")
fish_greedy <- fish_netclu_greedy$clusters
colnames(fish_greedy)<- c("ID", "NET_GREEDY")

all_clusters <- dplyr::left_join(fish_kmeans, fish_pam, by = "ID")
all_clusters <- dplyr::left_join(all_clusters, fish_hieraclust, by = "ID")
all_clusters <- dplyr::left_join(all_clusters, fish_greedy, by = "ID")
```

We now convert this `data.frame` into a long-format `data.frame`.

```{r}
all_long <- tidyr::pivot_longer(data = all_clusters,
                                cols = dplyr::contains("_"),
                                names_to = "Algorithm",
                                values_to = "Clusters")
all_long <- as.data.frame(all_long)
```

We now add back the geometry as an extra column to make this object spatial.

```{r}
all_long_sf <- dplyr::left_join(all_long,
                                custom[, c("ID", "geometry")],
                                by = "ID")
all_long_sf <- sf::st_as_sf(all_long_sf)
```

Now that we have a long-format spatial `data.frame`, we can take advantage of
the facets implemented in `ggplot2`.

```{r}
ggplot(all_long_sf) +
  geom_sf(aes(color = Clusters, fill = Clusters)) +
  geom_sf(data = europe, fill = "gray80") +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  theme_void() +
  facet_wrap(~ Algorithm)
```
