---
title: "5.2 Visualization"
author: "Pierre Denelle, Boris Leroy and Maxime Lenormand"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
bibliography: '`r system.file("REFERENCES.bib", package="bioregion")`' 
csl: journal-of-biogeography.csl    
vignette: >
  %\VignetteIndexEntry{5. Visualization}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 6)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library("bioregion")
    library("ggplot2")
    library("dplyr")
    library("rnaturalearth")
    library("sf")
    library("tidyr")
  })
})

options(tinytex.verbose = TRUE)
```

In this vignette, we aim at illustrating how to plot the bioregions identified
with the diferent algorithms available in `bioregion`.
<br>
Using one of the dataset coming along with `bioregion`, we show three strategies
to plot your results.

# Data  
For this vignette, we rely on the dataset describing the distribution of 
vascular plants in the Mediterranean part of France. We first load the matrix
format of this dataset, computes the dissimilarity matrix out of it and also
load the data.frame format of the data.

```{r}
data(vegemat)
vegedissim <- dissimilarity(vegemat, metric = "all")
data(vegedf)
```

Since we aim at plotting the result, we also need the object `fishsf` linking
each site of the dataset to a geometry.

```{r}
data(vegesf)
```

We also import the world coastlines, available from the `rnaturalearth`
R package.

```{r}
world <- rnaturalearth::ne_coastline(returnclass = "sf", scale = "large")
# Align the CRS of both objects
vegesf <- st_transform(vegesf, crs = st_crs(world))
```

# Plots  
In this section, we show three ways to plot your results.
<br>

## map_clusters()  
The first possibility is to use the function `map_clusters()` from the package.
This function can directly provide a plot of each site colored according to
the cluster they belong to.
<br>
Let's take an example with a K-means clustering, with a number of clusters set
to 3.

```{r}
vege_nhclu_kmeans <- nhclu_kmeans(vegedissim, n_clust = 3, index = "Simpson")
```

`map_clusters()` function can now simply takes the object `fish_nhclu_kmeans`,
which is of `bioregion.clusters` class, and the spatial distribution of sites,
stored in `fishsf`.

```{r}
map_clusters(vege_nhclu_kmeans, geometry = vegesf, plot = TRUE)
```

## Custom plot  
If you want to customize yourself the plot and not simply rely on the default
option, `map_clusters()` gives you the possibility to extract each site as well
as its geometry and cluster number.  
For this purpose, you can set the arguments like in the chunk below:

```{r}
custom <- map_clusters(vege_nhclu_kmeans, geometry = vegesf,
                       write_clusters = TRUE, plot = FALSE)
custom

# Crop world coastlines to the extent of the sf object of interest
europe <- sf::st_crop(world, sf::st_bbox(custom))

# Plot
ggplot(custom) +
  geom_sf(aes(fill = K_3), show.legend = FALSE) +
  geom_sf(data = europe) +
  scale_fill_viridis_d() +
  labs(title = "Kmeans with 3 clusters") +
  theme_void()
```

## Plot with facets  
Finally, you can be interested in plotting several bioregionalisations at once.
For this purpose, we can build a single `data.frame` gathering all the
bioregions obtained from distinct algorithms and then take advantage of the
`facets` implemented in `ggplot2`.
<br>

We first compute a few more bioregionalisation on the same dataset using
other algorithms.
<br>

```{r}
# Hierarchical clustering
set.seed(1)
vege_hclu_hierarclust <- hclu_hierarclust(dissimilarity = vegedissim,
                                          index = names(vegedissim)[6],
                                          method = "mcquitty", n_clust = 3)
vege_hclu_hierarclust$cluster_info

# Walktrap network bioregionalization
vegesim <- dissimilarity_to_similarity(vegedissim)
set.seed(1)
vege_netclu_walktrap <- netclu_walktrap(vegesim,
                                        index = names(vegesim)[6])
vege_netclu_walktrap$cluster_info # 3
```

We can now make a single `data.frame` with an extra-column indicating the
algorithm used.

```{r}
vege_kmeans <- vege_nhclu_kmeans$clusters
colnames(vege_kmeans)<- c("ID", "NHCLU_KMEANS")
vege_hieraclust <- vege_hclu_hierarclust$clusters
colnames(vege_hieraclust)<- c("ID", "HCLU_HIERARCLUST")
vege_walktrap <- vege_netclu_walktrap$clusters
colnames(vege_walktrap)<- c("ID", "NETCLU_Walktrap")

all_clusters <- dplyr::left_join(vege_kmeans, vege_hieraclust, by = "ID")
all_clusters <- dplyr::left_join(all_clusters, vege_walktrap, by = "ID")
```

We now convert this `data.frame` into a long-format `data.frame`.

```{r}
all_long <- tidyr::pivot_longer(data = all_clusters,
                                cols = dplyr::contains("_"),
                                names_to = "Algorithm",
                                values_to = "Clusters")
all_long <- as.data.frame(all_long)
```

We now add back the geometry as an extra column to make this object spatial.

```{r}
all_long_sf <- dplyr::left_join(all_long,
                                vegesf[, c("ID", "geometry")],
                                by = "ID")
all_long_sf <- sf::st_as_sf(all_long_sf)
```

Now that we have a long-format spatial `data.frame`, we can take advantage of
the facets implemented in `ggplot2`.

```{r}
ggplot(all_long_sf) +
  geom_sf(aes(color = Clusters, fill = Clusters)) +
  geom_sf(data = europe, fill = "gray80") +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  theme_void() +
  facet_wrap(~ Algorithm)
```

We can refine the above map by:
  * reordering the 3 bioregions so that they follow the same order
  * add some background for the Mediterranean sea and the mainland
  * crop the cells by the mainland
  * adjust labels
  
```{r}
world_countries <- rnaturalearth::ne_countries(scale = "large",
                                               returnclass = "sf")

# Background box
xmin <- st_bbox(world)[["xmin"]]; xmax <- st_bbox(world)[["xmax"]]
ymin <- st_bbox(world)[["ymin"]]; ymax <- st_bbox(world)[["ymax"]]
bb <- sf::st_union(sf::st_make_grid(st_bbox(c(xmin = xmin,
                                              xmax = xmax,
                                              ymax = ymax,
                                              ymin = ymin),
                                            crs = st_crs(4326)),
                                    n = 100))

# Crop world coastlines to the extent of the sf object of interest
vegesf <- st_transform(vegesf, crs = st_crs(world))
larger_bbox <- sf::st_bbox(vegesf)
larger_bbox[[1]] <- 1.5
larger_bbox[[2]] <- 42.15
larger_bbox[[3]] <- 8.1
larger_bbox[[4]] <- 45.3

europe <- sf::st_crop(world, larger_bbox)
sf_use_s2(FALSE)
europe_countries <- sf::st_crop(world_countries, larger_bbox)
europe_bb <- sf::st_crop(bb, larger_bbox)

plot_basis <- ggplot(europe) +
  geom_sf(data = europe_bb, fill = "aliceblue") +
  geom_sf(data = europe_countries, fill = "ivory", color = "gray50") +
  theme_void()

# Reordering bioregions
all_long_sf$bioregion <- all_long_sf$Clusters

all_long_sf[which(all_long_sf$Algorithm == "HCLU_HIERARCLUST" &
                    all_long_sf$Clusters == "2"), ]$bioregion <- "1"
all_long_sf[which(all_long_sf$Algorithm == "HCLU_HIERARCLUST" &
                    all_long_sf$Clusters == "1"), ]$bioregion <- "2"

all_long_sf[which(all_long_sf$Algorithm == "NHCLU_KMEANS" &
                    all_long_sf$Clusters == "3"), ]$bioregion <- "2"
all_long_sf[which(all_long_sf$Algorithm == "NHCLU_KMEANS" &
                    all_long_sf$Clusters == "1"), ]$bioregion <- "1"
all_long_sf[which(all_long_sf$Algorithm == "NHCLU_KMEANS" &
                    all_long_sf$Clusters == "2"), ]$bioregion <- "3"

all_long_sf[which(all_long_sf$Algorithm == "NETCLU_Walktrap" &
                    all_long_sf$Clusters == "3"), ]$bioregion <- "1"
all_long_sf[which(all_long_sf$Algorithm == "NETCLU_Walktrap" &
                    all_long_sf$Clusters == "1"), ]$bioregion <- "2"
all_long_sf[which(all_long_sf$Algorithm == "NETCLU_Walktrap" &
                    all_long_sf$Clusters == "2"), ]$bioregion <- "3"

# More readable labels for algorithms
all_long_sf$Algo <-
  ifelse(all_long_sf$Algorithm == "NHCLU_KMEANS", "K-Means",
         ifelse(all_long_sf$Algorithm == "NHCLU_PAM", "PAM",
                ifelse(all_long_sf$Algorithm == "HCLU_HIERARCLUST",
                       "Hierarchical Clustering", "Walktrap")))

# Cropping with borders of France
all_long_sf_france <-
  st_intersection(all_long_sf,
                  europe_countries[which(europe_countries$sovereignt == "France"), ])

# Plot
final_plot <- plot_basis +
  geom_sf(data = all_long_sf_france,
          aes(color = bioregion, fill = bioregion), show.legend = TRUE) +
  geom_sf(data = st_union(all_long_sf_france), fill = "NA", color = "gray50") +
  geom_sf(data = europe, fill = "gray50", linewidth = 0.1) +
  scale_color_viridis_d("Bioregion", option = "magma", direction = -1) +
  scale_fill_viridis_d("Bioregion", option = "magma", direction = -1) +
  theme_void() +
  theme(legend.position = "bottom") +
  facet_wrap(~ Algo)
final_plot
```
