---
title: "Microbenchmark"
author: "Maxime Lenormand, Boris Leroy and Pierre Denelle"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Microbenchmark}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 8)
# Packages --------------------------------------------------------------------
suppressPackageStartupMessages({
  suppressWarnings({
    library(bioRgeo)
    library(ade4)
    library(microbenchmark)
    library(ggplot2)
    library(tidyr)
    library(dplyr)
  })
})

options(tinytex.verbose = TRUE)

```

In this vignette, we aim at showing the differences in terms of computation
time between the different algorithms available in `bioRgeo`. For this purpose,
we use three datasets of varying size.
<br>
For each dataset, we use a site x species matrix as well as a dissimilarity
matrix which we compute using `bioRgeo::dissimilarity()`.

# 1. Datasets
## 1.1. Vegemat
The first dataset comes available with `bioRgeo` and was analyzed in
[this article](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.4718).
It contains the abundance of 3,697 plant species distributed in 715 sites in
the French Mediterranean area. 
From this dataset we use the two following files:
[vegedf](https://biorgeo.github.io/bioRgeo/reference/vegedf.html) a 
`data.frame` with 460,878 rows and 3 columns (Site, Species and Abundance) and
[vegemat](https://biorgeo.github.io/bioRgeo/reference/vegemat.html) 
a co-occurrence `matrix` containing the same information gathered in a `matrix`
with 715 rows and 3,697 columns.

```{r}
data(vegedf)
data(vegemat)

vegemat_dissim <- dissimilarity(vegemat,  metric = c("abc", "Simpson"))
vegemat_df <- mat_to_net(vegemat, weight = TRUE, remove_zeroes = TRUE)
```

## 1.2. Benthic invertebrates
The second dataset also comes available with `bioRgeo` and contains
distribution of 838 invertebrate species in the Bretagne region, France.

```{r}
contin <- t(as.matrix(readRDS("../data-raw/benthic_invertebrates.RDS")))

contin_dissim <- dissimilarity(contin, metric = "all")
contin_df <- mat_to_net(contin, weight = TRUE, remove_zeroes = TRUE)
```

## 1.3. Aravo
The third dataset, called `aravo`, is retrieved from the package `ade4`. It
contains the distribution of 82 alpine plants in 75 sites distributed in the
French Alps.

```{r}
data("aravo")
aravo <- as.matrix(aravo$spe)
dim(aravo)

ara_dissim <- dissimilarity(aravo, metric = "all")

ara_df <- mat_to_net(aravo, weight = TRUE, remove_zeroes = TRUE)
```

# 2. Microbenchmark
We here assessing the time needed by each clustering algorithm on the three
datasets loaded above.

```{r}
mbm <- suppressMessages(
  microbenchmark(
    # aravo
    ara_nhclu_dbscan = nhclu_dbscan(dissimilarity = ara_dissim,
                                    index = "Simpson", plot = FALSE),
    ara_nhclu_kmeans = nhclu_kmeans(ara_dissim, n_clust = 2:10,
                                    index = "Simpson"),
    ara_nhclu_pam = nhclu_pam(ara_dissim, n_clust = 2:10, index = "Simpson"),
    ara_hclu_hierarclust = hclu_hierarclust(dissimilarity = ara_dissim,
                                            n_clust = 5),
    ara_hclu_optics = hclu_optics(ara_dissim, index = "Simpson"),
    ara_netclu_beckett = netclu_beckett(ara_df),
    ara_netclu_greedy = netclu_greedy(ara_df),
    ara_netclu_labelprop = netclu_labelprop(ara_df),
    ara_netclu_leadingeigen = netclu_leadingeigen(ara_df),
    ara_netclu_oslom = netclu_oslom(ara_df),
    ara_netclu_walktrap = netclu_walktrap(ara_df),
    ara_netclu_infomap = netclu_infomap(ara_df),
    ara_netclu_louvain = netclu_louvain(ara_df),
    
    # benthic vertebrates
    benthic_nhclu_dbscan = nhclu_dbscan(dissimilarity = contin_dissim,
                                        index = "Simpson", plot = FALSE),
    benthic_nhclu_kmeans = nhclu_kmeans(contin_dissim, n_clust = 2:10,
                                        index = "Simpson"),
    benthic_nhclu_pam = nhclu_pam(contin_dissim, n_clust = 2:10, index = "Simpson"),
    benthic_hclu_hierarclust = hclu_hierarclust(dissimilarity = contin_dissim,
                                                n_clust = 5),
    benthic_hclu_optics = hclu_optics(contin_dissim, index = "Simpson"),
    benthic_netclu_beckett = netclu_beckett(contin_df),
    benthic_netclu_greedy = netclu_greedy(contin_df),
    benthic_netclu_labelprop = netclu_labelprop(contin_df),
    benthic_netclu_leadingeigen = netclu_leadingeigen(contin_df),
    benthic_netclu_oslom = netclu_oslom(contin_df),
    benthic_netclu_walktrap = netclu_walktrap(contin_df),
    benthic_netclu_infomap = netclu_infomap(contin_df),
    benthic_netclu_louvain = netclu_louvain(contin_df),
    
    # vegetation
    veg_nhclu_dbscan = nhclu_dbscan(dissimilarity = vegemat_dissim,
                                    index = "Simpson", plot = FALSE),
    veg_nhclu_kmeans = nhclu_kmeans(vegemat_dissim, n_clust = 2:10,
                                    index = "Simpson"),
    veg_nhclu_pam = nhclu_pam(vegemat_dissim, n_clust = 2:10,
                              index = "Simpson"),
    veg_hclu_hierarclust = hclu_hierarclust(dissimilarity = vegemat_dissim,
                                            n_clust = 5),
    veg_hclu_optics = hclu_optics(vegemat_dissim, index = "Simpson"),
    # veg_netclu_beckett = netclu_beckett(vegemat_df),
    veg_netclu_greedy = netclu_greedy(vegemat_df),
    veg_netclu_labelprop = netclu_labelprop(vegemat_df),
    veg_netclu_leadingeigen = netclu_leadingeigen(vegemat_df),
    # veg_netclu_oslom = netclu_oslom(vegemat_df),
    veg_netclu_walktrap = netclu_walktrap(vegemat_df),
    veg_netclu_infomap = netclu_infomap(vegemat_df),
    veg_netclu_louvain = netclu_louvain(vegemat_df),
    
    times = 2))

mbm_plot <- data.frame(mbm)
mbm_plot$expr <- as.character(mbm_plot$expr)
mbm_plot$dataset <- ifelse(
  grepl("ara_", mbm_plot$expr),
  paste0("Aravo (", nrow(aravo) * ncol(aravo), " cells)"),
  ifelse(grepl("benthic_", mbm_plot$expr),
         paste0("Benthic (", nrow(contin) * ncol(contin), " cells)"),
         paste0("Vegetation (", nrow(vegemat) * ncol(vegemat), " cells)")))

mbm_plot$algorithm <- gsub( ".*_", "", mbm_plot$expr)

# Time in minutes
mbm_plot$time_min <- mbm_plot$time / 60e9
# Time in seconds
mbm_plot$time_sec <- mbm_plot$time / 1e9
```

Plotting the results.

```{r, eval = FALSE, echo = FALSE}
label_plot <- data.frame(
  algorithm = rep(unique(mbm_plot$algorithm), length(unique(mbm_plot$dataset))),
  dataset = rep(unique(mbm_plot$dataset), length(unique(mbm_plot$algorithm))))

labeldat <- mbm_plot %>%
  group_by(dataset, algorithm) %>%
  summarize(ypos = max(time) + 0.25 ) %>%
  inner_join(., label_plot) %>%
  as.data.frame()

ggplot(mbm_plot, aes(as.factor(dataset), time)) +
  # geom_boxplot(aes(fill = reorder(algorithm, time)), show.legend = FALSE) +
  geom_boxplot(aes(color = algorithm, fill = algorithm), show.legend = FALSE) +
  geom_text(data = labeldat, aes(x = as.factor(dataset), y = ypos, label = algorithm,
                                 color = algorithm), show.legend = FALSE) +
  scale_color_viridis_d("Algorithm") +
  scale_fill_viridis_d("Algorithm") +
  labs(x = "Dataset", y = "Time (milliseconds)") +
  theme_classic() +
  theme(panel.border = element_rect(fill = NA))
```


```{r}
ggplot(mbm_plot[which(mbm_plot$dataset == "Aravo (6150 cells)"), ],
       aes(reorder(algorithm, time_sec), time_sec)) +
  geom_boxplot(aes(color = reorder(algorithm, time_sec),
                   fill = reorder(algorithm, time_sec)), show.legend = FALSE) +
  scale_color_viridis_d("Algorithm") +
  scale_fill_viridis_d("Algorithm") +
  labs(title = "Aravo (6,510 cells)", x = "", y = "Time (seconds)") +
  theme_classic() +
  theme(panel.border = element_rect(fill = NA),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```


```{r}
ggplot(mbm_plot[which(mbm_plot$dataset == "Benthic (135756 cells)"), ],
       aes(reorder(algorithm, time_sec), time_sec)) +
  geom_boxplot(aes(color = reorder(algorithm, time_sec),
                   fill = reorder(algorithm, time_sec)), show.legend = FALSE) +
  scale_color_viridis_d("Algorithm") +
  scale_fill_viridis_d("Algorithm") +
  labs(title = "Benthic invertebrates (135,756 cells)",
       x = "", y = "Time (seconds)") +
  theme_classic() +
  theme(panel.border = element_rect(fill = NA),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

OSLOM and Beckett too slow, not ran so far on the Vegetation data.

```{r}
ggplot(mbm_plot[which(mbm_plot$dataset == "Vegetation (2643355 cells)"), ],
       aes(reorder(algorithm, time_min), time_min)) +
  geom_boxplot(aes(color = reorder(algorithm, time_min),
                   fill = reorder(algorithm, time_min)), show.legend = FALSE) +
  scale_color_viridis_d("Algorithm") +
  scale_fill_viridis_d("Algorithm") +
  labs(title = "Vegetation (2,643,355 cells)", x = "", y = "Time (minutes)") +
  theme_classic() +
  theme(panel.border = element_rect(fill = NA),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```
